Intrinsic Verification of Parsers and Formal Grammar Theory in Dependent Lambek Calculus

### 2025-03-23 paper/paper.tex
need to write up non linear inductive rules

\begin{figure}
  \begin{tikzpicture}[node distance = 25mm ]
    \node[state, initial] (0) {$\0$};
    \node[state, below left of=0] (1) {$\1$};
    \node[state, right of=1, accepting] (2) {$\2$};

    \path[->] (0) edge[above] node{$\epsilon$} (1)
              (0) edge[right] node{$\stringquote{c}$} (2)
              (1) edge[loop left] node{$\stringquote{a}$} (1)
              (1) edge[below] node{$\stringquote{b}$} (2);
  \end{tikzpicture}
  \begin{lstlisting}
data Trace : (s : Fin 3) (*@$\to$@*) L where
  stop : (*@$\uparrow$@*)(Trace 2)
  1to1 : (*@$\uparrow$@*)('a' (*@$\lto$@*) Trace 1 (*@$\lto$@*) Trace 1)
  1to2 : (*@$\uparrow$@*)('b' (*@$\lto$@*) Trace 2 (*@$\lto$@*) Trace 1)
  0to2 : (*@$\uparrow$@*)('c' (*@$\lto$@*) Trace 2 (*@$\lto$@*) Trace 0)
  0to1 : (*@$\uparrow$@*)(Trace 1 (*@$\lto$@*) Trace 0)

@@

\end{lstlisting}
  \caption{NFA for $(\a^{*} \otimes \b) \oplus \c$ and its corresponding type}
  \label{fig:exampleNFA}
\end{figure}

@@

@@

    \inferrule{~}{\ctxwffjdg \Gamma {Nat}}

@@

\subsection{Non-linear Typing}
The non-linear types of \theoryabbv include the standard dependent
types for $\Pi$,$\Sigma$, extensional equality, natural numbers,
\steven{,just take Booleans to be implemented via $1$ and $\Sigma$,} unit, and empty types, and we assume function
extensionality\cite{Hofmann_1997}. We present a selection of
the non-linear type constructions in
\Cref{fig:non-linear-types} and provide the rest in
\cref{fig:full-non-linear-types, fig:jdg-eq-nonlinear} of the appendix.
First, we include universe types $U$ of
small non-linear types and $L$ of linear types. These are defined as
universes ``ala Coquand'' in that we define a judgment saying when a
non-linear type is small and define the universe to internalize
precisely this judgment \cite{coquandPresheafModel,lmcs:7713}. These universe
types are needed so that we can define types by recursion on natural
numbers. As in \cite{krishnaswami_integrating_2015}, the extensional equality type
$M =_{X} N$ used here has the introduction form $refl$, but no
elimination due to the presence of equality reflection. Next, we include a non-linear type $\ltonl A$ where $A$ is a
linear type. The intuition for this type is that its elements are the
linear terms that are ``resource free'': its introduction rule says we
can construct an $\ltonl A$ when we have a linear term of type $A$
with no free linear variables. This type is used extensively in our
examples, playing a similar role to the $!$ modality of ordinary
linear logic or the persistence modality $\square$ of separation logic
\cite{girard_linear_1987,jung_higher-order_2016}.

@@

    \inferrule{~}{\nonlinterm {\Gamma} {0} {Nat}}

    \inferrule{\nonlinterm {\Gamma} {n} {Nat}}{\nonlinterm {\Gamma} {suc~n} {Nat}}

    \inferrule{\ctxwffjdg {\Gamma , n : Nat} {X(n)} \and
           \nonlinterm {\Gamma} {M_0} {X(0)} \and
           \nonlinterm {\Gamma, n : Nat, x : X(n)} {M_{suc}(n , x)} {X(suc~n)}}
          {\nonlinterm {\Gamma} {elim_{Nat}(X,M_0,M_{suc})(n)} {X(n)}}

@@

\paragraph{\cref{ax:string-top}}
\begin{theorem}
  In $\Grammar$, $\semg{!}$ has an inverse.
\end{theorem}
\begin{proof}
  Because $\semgg{\top}\gamma\,w$ is a singleton set, it suffices to show that
  $\semg{\StringGram}\gamma\,w$ is likewise a singleton set for all $\gamma$ and
  $w$.

>>>>

\begin{figure}
  \begin{minipage}[t]{.6\textwidth}
    \vspace{0pt}
  \begin{lstlisting}
data Trace : (s : Fin 3) (*@$\to$@*) L where
  stop : (*@$\uparrow$@*)(Trace 2)
  1to1 : (*@$\uparrow$@*)('a' (*@$\lto$@*) Trace 1 (*@$\lto$@*) Trace 1)
  1to2 : (*@$\uparrow$@*)('b' (*@$\lto$@*) Trace 2 (*@$\lto$@*) Trace 1)
  0to2 : (*@$\uparrow$@*)('c' (*@$\lto$@*) Trace 2 (*@$\lto$@*) Trace 0)
  0to1 : (*@$\uparrow$@*)(Trace 1 (*@$\lto$@*) Trace 0)

@@

\end{lstlisting}
  \end{minipage}%
  \begin{minipage}[t]{.4\textwidth}
  \vspace{0pt}
  \hspace{10pt}
  \begin{tikzpicture}[node distance = 25mm ]
    \node[state, initial] (0) {$\0$};
    \node[state, below left of=0] (1) {$\1$};
    \node[state, right of=1, accepting] (2) {$\2$};

    \path[->] (0) edge[above] node{$\epsilon$} (1)
              (0) edge[right] node{$\stringquote{c}$} (2)
              (1) edge[loop left] node{$\stringquote{a}$} (1)
              (1) edge[below] node{$\stringquote{b}$} (2);
  \end{tikzpicture}
  \end{minipage}%
  \caption{NFA for $(\a^{*} \otimes \b) \oplus \c$ and its corresponding type}
  \label{fig:exampleNFA}
\end{figure}

@@

\steven{TODO define general inductives as the least fixed point of strictly
  positive endofunctors on $U$}

\steven{TODO linear $\mu$ in the type formation rules? nonlinear $\mu$ in the
  type formation rules?}

@@

    % \inferrule{~}{\ctxwffjdg \Gamma {Nat}}

@@

\subsection{Non-linear Typing}
The non-linear types of \theoryabbv include the standard dependent
types for $\Pi$,$\Sigma$, extensional equality, unit and empty types, and
inductive types that are defined strictly positively. We may also assume function
extensionality\cite{Hofmann_1997}; Function
extensionality is holds in Cubical Agda, and in our implementation we make heavy use of it of when
validating that the denotational semantics respects the equational theory of
\theoryabbv. However, it is not necessary to assume function extensionality for
the non-linear types in \theoryabbv, as we don't utilize it in any of our constructions.

We present a selection of
the non-linear type constructions in
\Cref{fig:non-linear-types} and provide the rest in
\cref{fig:full-non-linear-types, fig:jdg-eq-nonlinear} of the appendix.
First, we include universe types $U$ of
small non-linear types and $L$ of linear types. These are defined as
universes ``ala Coquand'' in that we define a judgment saying when a
non-linear type is small and define the universe to internalize
precisely this judgment \cite{coquandPresheafModel,lmcs:7713}. These universe
types are needed so that we can define types by recursion on natural
numbers. As in \cite{krishnaswami_integrating_2015}, the extensional equality type
$M =_{X} N$ used here has the introduction form $refl$, but no
elimination due to the presence of equality reflection. Next, we include a non-linear type $\ltonl A$ where $A$ is a
linear type. The intuition for this type is that its elements are the
linear terms that are ``resource free'': its introduction rule says we
can construct an $\ltonl A$ when we have a linear term of type $A$
with no free linear variables. This type is used extensively in our
examples, playing a similar role to the $!$ modality of ordinary
linear logic or the persistence modality $\square$ of separation logic
\cite{girard_linear_1987,jung_higher-order_2016}.

@@

    % \inferrule{~}{\nonlinterm {\Gamma} {0} {Nat}}

    % \inferrule{\nonlinterm {\Gamma} {n} {Nat}}{\nonlinterm {\Gamma} {suc~n} {Nat}}

    % \inferrule{\ctxwffjdg {\Gamma , n : Nat} {X(n)} \and
    %        \nonlinterm {\Gamma} {M_0} {X(0)} \and
    %        \nonlinterm {\Gamma, n : Nat, x : X(n)} {M_{suc}(n , x)} {X(suc~n)}}
    %       {\nonlinterm {\Gamma} {elim_{Nat}(X,M_0,M_{suc})(n)} {X(n)}}

@@

\paragraph{\cref{ax:string-top}}
\begin{theorem}
  In $\Grammar$, $\semg{!}$ has an inverse.
\end{theorem}
\begin{proof}
  Because $\semg{\top}\gamma\,w$ is a singleton set, it suffices to show that
  $\semg{\StringGram}\gamma\,w$ is likewise a singleton set for all $\gamma$ and
  $w$.
